---
title: "Use-case: Investigating the Impact of Sampling Co-location"
output: 
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Use-case: Investigating the Impact of Sampling Co-location}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

In ecological studies, monitoring multiple species across various taxa (e.g., butterflies, plants, birds) is essential for understanding biodiversity and ecosystem health. One strategy for monitoring these species is co-location, where data collection for different taxa occurs at the same sites. This approach can have various advantages, such as cost-effectiveness and comprehensive data collection, but it may also introduce biases or limitations.

This vignette demonstrates a workflow to simulate the effects of co-location on species monitoring data using the STRIDER package. We will generate species presence-absence data across a number of sites and then simulate different scenarios of co-located and non-co-located sampling. This includes:

 * Generating Full Sampling Data: Creating a full dataset with species counts and environmental data across multiple sites.
 * Simulating Co-location: Using a custom function to thin the dataset, representing different degrees of co-location for monitoring different taxa.
 * Visualizing Co-location Effects: Plotting the distribution of species monitoring data to understand how co-location impacts data representation.
 
By running these simulations, we aim to investigate the benefits and challenges of co-location, such as whether certain taxa benefit from co-located monitoring and how sampling strategies might influence the accuracy and reliability of biodiversity assessments.

## Setup

First, we load the required packages and set a seed for reproducibility.

```{r setup}
library(STRIDER)
library(terra)
library(sf)
library(dplyr)
library(virtualspecies)
library(ggplot2)
set.seed(42)
```

## Custom Environmental State

We define custom environmental variables with different spatial patterns to simulate realistic scenarios.

```{r custom environmental state}
dim_x <- 500
dim_y <- 500

#elevation
e_elev <- matrix(runif(dim_x * dim_y), dim_x, dim_y) |> 
  rast() |>
  terra::aggregate(fact = 10) |> 
  terra::disagg(fact = 10) |> 
  focal(15,expand=T,fun = "mean") |> 
  focal(15,expand=T,fun = "mean") *1000

#slope and aspect
e_slope <- e_elev |> terrain()
e_slope[is.na(e_slope)] <- mean(values(e_slope),na.rm = T)
e_aspect <- e_elev |> terrain(v="aspect")
e_aspect[is.na(e_aspect)] <- mean(values(e_aspect),na.rm = T)

#latitudinal gradient, - elevation
e_temperature <- 60 + terra::rast(matrix(rep(seq(from = 1, to = dim_x), times = dim_y), dim_x, dim_y))/100 - (e_elev/100)*10

#latitudinal gradient, - elevation
e_rainfall <- 20 + t(terra::rast(matrix(rep(seq(from = 1, to = dim_x), times = dim_y), dim_x, dim_y)))/100 + (e_elev/100)*10
e_rainfall <- e_rainfall/max(values(e_rainfall))


custom_env <- c(e_elev,e_slope,e_aspect,e_temperature,e_rainfall)

# Define environmental variables with different spatial patterns

names(custom_env) <- c("elevation", "slope", "aspect", "temperature", "rainfall")
```

Next, we integrate these custom environmental variables into the simulation object.

```{r integrate custom environment}
background <- terra::rast(matrix(0, dim_x, dim_y))
sim_obj <- SimulationObject(background = background)
sim_obj <- sim_state_env(sim_obj, spatraster = custom_env)
```

## Custom Suitability Functions for Multiple Species

We define custom suitability functions for multiple species, each influenced differently by the environmental variables. We use the `virtualspecies` package to do this.

```{r custom suitability functions}

# virtual species
suitability_virtual_species <- function(simulation_object,n_targets = 1,...){
  outs <- list()
  for (i in 1:n_targets){
    out <- virtualspecies::generateSpFromPCA(simulation_object@state_env,sample.points = T, nb.point = 300,plot = F,...)
    outs[[i]] <- out$suitab.raster
  }
  names(outs) <- paste0("target_",1:n_targets)
  rast(outs)
}

sim_obj <- sim_state_target_suitability(sim_obj, fun = suitability_virtual_species,n_targets = 5)
```

Realised suitability

```{r realise}
sim_obj <- sim_state_target_realise(sim_obj,fun=state_target_realise_binomial)
```

## Custom Effort Function

We create a custom effort function that simulates sampling effort.

```{r custom effort function}
custom_effort_function <- function(sim_obj, n_sites) {
  sites <- sample(cells(sim_obj@background), n_sites, replace = TRUE)
  coords <- xyFromCell(sim_obj@background, sites)
  effort_df <- data.frame(
    x = coords[, 1],
    y = coords[, 2]
  )
  effort_sf <- st_as_sf(effort_df, coords = c("x", "y"))
  
  return(effort_sf)
}

sim_obj <- sim_effort(sim_obj, fun = custom_effort_function, n_sites = 100)
```

## Custom Detection and Reporting Functions

We could define custom detection and reporting functions to introduce variability in the detection and reporting probabilities, but just use the default for now.

```{r custom detection and reporting functions}
#detection
sim_obj <- sim_detect(sim_obj,fun = detect_equal, prob = 1)

# Simulate the reporting
sim_obj <- sim_report(sim_obj,fun = report_equal, prob = 1, platform = "iRecord")
```

## Visualisation

Visualised the simulation objects

```{r visualisation}
plot(sim_obj)
```

## Investigating co-location

This code demonstrates how to simulate co-location of species monitoring data at various sites and visualize the distribution of targets (species) across these sites. The goal is to investigate the effects of co-locating monitoring efforts for different taxa versus individual monitoring efforts. This is achieved through a function `colocation_thinning()` that thins the data to simulate varying degrees of co-location.

```{r colocation functions}
# Export the full sampling data from the simulation object
df <- sim_obj |> export_df()

# Get the full number of samples
nrow(df)

# Function to thin rows to simulate colocation
colocation_thinning <- function(sim_df, colocate_rate = 1, consistent_colocation = F, n_samples = 100) {
  # Determine the number of unique targets (species) in the dataset
  n_targets <- length(unique(sim_df$target))
  
  # Calculate the number of targets per site based on the colocation rate
  targets_per_site <- round(colocate_rate * n_targets)
  print(paste0(targets_per_site, " targets per site"))
  
  if (consistent_colocation) {
    # If consistent colocation is required, sample a subset of targets to colocate
    colocated_targets <- sample(unique(sim_df$target), targets_per_site)
    
    # Filter the data for the colocated targets
    sim_df_colocated <- df |> filter(target %in% colocated_targets)
    
    # Randomly select a subset of sites to colocate the targets
    sites_to_select <- unique(df$ID) |> sample(round(n_samples * colocate_rate) / targets_per_site)
    sim_df_colocated <- sim_df_colocated |> filter(ID %in% sites_to_select)
    
    # Sample the remaining targets to maintain the desired sample size
    sim_df_noncolocated <- sim_df %>%
      filter(!(target %in% colocated_targets)) %>%
      sample_n(round(n_samples * (1 - colocate_rate)))
    
    # Combine the colocated and non-colocated samples
    sim_df <- bind_rows(sim_df_colocated, sim_df_noncolocated)
    
  } else {
    # Sample a subset of targets for each site if colocation is not consistent
    sim_df <- df |> group_by(ID) |> sample_n(targets_per_site)
    
    # Thin the sites to match the required number of samples
    sites_to_select <- unique(df$ID) |> sample(round(n_samples / targets_per_site))
    sim_df <- sim_df |> filter(ID %in% sites_to_select)
  }

  sim_df
}

plot_colocation <- function(thinned_df){
  thinned_df |>
    ggplot(aes(x = as.factor(ID), y = target, colour = target)) +
    geom_point() +
    theme_minimal() +
    xlab("Site")
}

```

### Here are some different co-location scenarios

Despite different co-location, each scenario has consistent effort: 100 samples (the function default).

In this scenario, all targets are sampled at every site. This represents full co-location where each monitoring site collects data on all taxa.

```{r scen1}
colocation_thinning(df, colocate_rate = 1) |> plot_colocation()
```

In this scenario, 60% of the targets are sampled at each site. The specific targets that are sampled can vary from site to site, leading to inconsistent co-location. Different sets of 3 out of 5 targets are sampled at each site.

```{r scen2}
colocation_thinning(df, colocate_rate = 0.6) |> plot_colocation()
```

Here, only 20% of the targets are sampled at each site, which means each site samples only a single taxa with no overlap. This represents minimal co-location.

```{r scen3}
colocation_thinning(df, colocate_rate = 0.2) |> plot_colocation()
```

In this scenario, 60% of the targets are consistently sampled at the same sites. The same 3 out of 5 targets are always sampled at specific sites, while the remaining 2 targets are sampled randomly. This ensures some level of consistent co-location for certain taxa.

```{r scen4}
colocation_thinning(df, colocate_rate = 0.6,consistent_colocation = T) |> plot_colocation()
```

## Data analysis

The resulting data frame from each co-location scenario could be used to test various hypotheses around the advantages/disadvantages of sampling co-location.

```{r scen4 data}
colocation_thinning(df, colocate_rate = 0.6,consistent_colocation = T)
```






